Visualizing Independence Using Extended Association Plots

### 2005-08-03 trunk/vcd/paper/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@428 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

However, it has two major drawbacks: First,
the significance level for the hard-coded critical values 2 and 4 is
usually unknown.

  - for non-significant results, clearly reduce the amount of color, i.e. the maximum
    chroma used.

In particular,
the Friendly shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
as in general it is unknown to which significance level $\alpha$ the
values $c_\alpha = 2$ and $4$ correspond for the table being visualized.

As pointed out in Section \ref{sec:tests}, the critical values for
the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
from the permutation distribution. Instead of the hard-coded
values 2 and 4, the particular critical values, e.g., at the levels
$\alpha = 0.1$ and $0.01$, for the table to be visualized could be
used. Hence, exactly those residuals causing the (potential)
dependency within the table are highlighted. At the moment, these critical values
are derived in \texttt{vcd} by simulation of the underlying distribution.

To illustrate what a `postive' example looks like we give the association plot
for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
color and gender of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
shows the association plot for the 328 female students. It is easy to see that
the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
due to higher frequencies of students with hair and eye color
black/brown and blond/blue and lower frequencies of
blond/brown and brown/blue than would be plausible under independence. Furthermore,
it can also be seen that the residual corresponding to the black/blue
cell is significant at the 10\% level.

>>>>

The shading scheme of \cite{vcd:Friendly:1994} was suggested to visualize
the pattern of dependence in contingency tables, as discussed above, but
the presence (or absence) of colors in a plot also always conveys an
impression of interestingness (or uninterstingness respectively). That is,
statisticians might be tempted to interpret the absence of color in a plot
as a clue that there is no significant departure from independence. Or vice
versa, colored cells would convey the impression that there is significant
dependence. Currently, both is not true as can be seen in the left panel
of Figure~\ref{fig:shadings} which shows the mosaic display for the arthritis
data with Friendly shading. Although there is significant dependence, no 
residual exceeds an absolute value of $2$ and hence no cell is colored.
Of course, it can be argued that the shading was not designed for this purpose
and that different cut offs than $2$ and $4$ should be used here. But then
again, it would be nice if these cut offs could be chosen automatically in a
data-driven fashion which is what we do below.

The Friendly shading can be interpreted to be a visualization of the maximum
statistic $M$ (\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
of $2$ and $4$ respectively. But the problem is that it is not clear to which
significance levels $\alpha$ these critical values correspond because the distribution
of $M$ depends on the underlying contingency table. The natural solution to this
problem is to compute the critical values from the distribution of $M$ in a data-driven
way (i.e., for the table visualized) and use these instead of the hard-coded values
$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
the arthritis data by employing the critical values
$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
(and using the diverging HCL palette). By using these cut offs, the presence of color
in the plot is equivalent to significance (of the maximum statistic $M$) at 
level $\alpha = 0.1$ and $\alpha = 0.01$ respectively and exactly the cells which `cause'
the dependence are highlighted. For the arthritis data, these are in particular the 
cells in the last column that signal that there are significantly more
treated patients and fewer placebo patients with marked improvement than would be expected
under independence.

The significance $alpha = 0.1$ and $0.01$ are chosen because this leads to displays where
fully colored cells are clearly significant (i.e., with a $p$ value below $0.01$),
cells without color are clearly non-significant (i.e., with a $p$ value above $0.1$), and
cells in between can be considered to be weakly significant (i.e., $0.1 \le p < 0.01$).
Of course, users could choose any other set of significance levels they feel comfortable
with, e.g., only a single cutoff at $\alpha = 0.05$ or three cut offs at $0.1$, $0.05$ and $0.01$
etc. Another option could be to use a continuous shading where the $p$ value corresponding
to a cell controls the interpolation between the neutral and the full color. However, our
experience is that this typically results in too much color in a plot which in turn tends
to conceal the important cells and over-emphasizes the unimportant ones. Hence, we stick
to a shading with fewer colors that are easier to interpret.

This extended shading is already very flexible and combines visualization and inference,
but it should only be applied when using the maximum statistic is appropriate. For 
the sum-of-squares statistic $X^2$ (and all other functionals $\lambda(\cdot)$), this
shading cannot be used.

use pre-determined (or manually adjusted) cut offs, but visualize the significance
by the amount of color in the plot, i.e., use only little color if the corresponding
test result is non-significant.

For HCL, the amount of color can easily be controlled by varying the maximum
chroma value. So far $100$, now $20$.

For HSV, a similar effect can be achieved by varying the value dimension which
was not used up to now. Setting value to $v = 0.5$ a darker palette of colors can be
obtained.
